---
title: "Introduction to Computer Programming with R (FOR 6934)"
author: 
- "Qing Zhao (School of Forest Resources & Conservation, qing.zhao@ufl.edu)"
- "Daijiang Li (Department of Wildlife Ecology & Conservation, dli1@ufl.edu)"
- "Denis Valle (School of Forest Resources & Conservation, drvalle@ufl.edu)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: false
    css: myStyle.css
---

## Basic Information
- Self introduction
- Why use R
    - Free
    - Powerful (and fast growing)
    - Programming skills are transferable
- Structure of lectures
- Homework and grading
- Office hour

## Learning outcomes
- Data type and index
- Loops
- Customized functions
- Figures & tables
- Enjoy computer programming

## Class one 
- help() and several other basic functions
- Understand data types and structures
- Convert data types and structures



## help() function
- You can learn to use all of the functions in R by using help() function
- You can also use the help() function to learn about data sets and packages
- Use it to explore the infinite potential of R


- Other <b>useful</b> functions
    - str()
    - apropos()

## Sample code
Use help() to understand functions
```{r, eval=F}
help(mean)
help(apply)
help(ifelse)
```

## Sample code
? is the same as help()
```{r, eval=F}
help(sum)
?sum
```

## Sample code
Use help() to understand control-flow constructs
```{r, eval=F}
help("if")
help("else")
help("for")
```

## Sample code
Use help() to understand data sets and packages
```{r, eval=F}
help(cars)
help(package="boot")
```

## Sample code
str() and apropos()
```{r}
str(mean)
apropos("mean")
```

## Data types
Data type           Example             
------------------- --------------------
Logical             TRUE (T), FALSE (F)  
Numeric (double)    4, 3.5, 1e2          
Integer             2L, 1e2L             
Complex             5+3i                 
Character           "Tom", "1.2", "FALSE" 

## Data structures
Data structure Explanation 
-------------- -----------
Vector         A set of values of a single type
Matrix         A two-dimensional rectangular data set of a single type of data
Array          A multiple dimensional data set of a single type of data
Data frame     A number of vectors of equal length (like a matrix). Each vector (column, also called variable) can have a different data type
List           An R object containing different types of data, functions, and even another list

## Sample code
Create vectors with logical values
```{r}
c(TRUE, TRUE, FALSE)
```

## Sample code
Create vectors with numeric & integer values
```{r}
c(5.2, 3.3, 4.9, 6.4, 8.1)
c(1L, 2L, 3L)
1:3
```

## Sample code
Create vectors with character values
```{r}
c("red", "blue", "pink")
c("3.2", "1.5", "7.3")
c("TRUE", "FALSE")
```

## Sample code
Vectors do not allow for mixed types of values
```{r}
c(FALSE, 1.52, -2L, TRUE)
c(FALSE, 1.52, -2L, TRUE, 'a')
```

## Sample code
Name vectors
```{r}
a <- c(3.52, 5.73, 4.83)
a

b = c(TRUE, FALSE, TRUE, FALSE)
b

c <- c("a", "b", "c") # '' also works, pick your style
c
```

## Sample code
Create and name matrices
```{r}
mat1 <- matrix(c("dog", "cat", "horse", "gator"), ncol=2, nrow=2, byrow=FALSE)
mat1
mat2 <- matrix(c("dog", "cat", "horse", "gator"), ncol=2, nrow=2, byrow=TRUE)
mat2
```

## Sample code
Create and name arrays
```{r}
arr <- array(1:30, dim=c(3,5,2))
arr
```

## Sample code
Create and name data frames
```{r}
animal <- c('dog', 'cat', 'horse', 'tortoise')
age <- c(15, 1, 3, 100)
large <- c(FALSE, FALSE, TRUE, TRUE)
dat <- data.frame(animal, age, large)
dat
```

## Sample code
Create and name lists
```{r}
animal <- c('dog', 'cat')
age <- c(15, 1)
dat <- data.frame(animal, age)
lst <- list(animal=animal, age=age, dat=dat)
lst
```

## Functions for identifying data types and structures
- class()
- typeof()
- str()

## Functions for specific data types
- is.logical()
- is.numeric()
- is.integer()
- is.character()

## Functions for specific data structures
- is.vector()
- is.matrix()
- is.array()
- is.data.frame()
- is.list()

## Sample code
Identify data types and structure in vectors
```{r}
a <- c(3.25, 6.73, 5.5)
class(a)
typeof(a)
str(a)
```

## Sample code
Identify specific data types in vectors
```{r}
is.numeric(a)
is.character(a)
```

## Sample code
Identify specific data structures in vectors
```{r}
is.vector(a)
is.matrix(a)
```

## Sample code
Identify data types and structures in matrices
```{r}
b <- matrix(c('dog', 'cat', 'horse', 'gator'), nrow=2, ncol=2)
class(b)
typeof(b)
str(b)
```

## Sample code
Identify specific data types in matrices
```{r}
is.character(b)
is.logical(b)
```

## Sample code
Identify specific data structures in matrices
```{r}
is.data.frame(b)
is.matrix(b)
```

## Sample code
Identify data structures in data frames and lists
```{r}
dat <- data.frame(animal=c('dog', 'cat'), age=c(15,1))
lst <- list(dat, fruit=c('apple', 'peach'))

class(dat)
class(lst)
```

## Some comments for data types and structures
- Always use class() to identify data structures first. It will give you the data type if it is a vector, or the data structure otherwise
- If it is a matrix or array, use typeof() to identify the data type in it
- typeof() does not make much sense for data frames and lists, since they allow multiple types of data
- Use class() and typeof() on the columns in a data frame or components in a list

## Functions for converting data types
- as.logical()
- as.numeric()
- as.integer()
- as.character()

## Functions for converting data structures
- as.vector()
- as.matrix()
- as.data.frame()

## Sample code
Convert data to logical values
```{r}
as.logical(c(1.5, -2.2, 1000, 0))
as.logical(c(1L, 0L, 5e2L, -3L))
as.logical(c('a', 'b', 'c'))
```

## Sample code
Convert data to numeric values
```{r}
as.numeric(c(TRUE, TRUE, FALSE, TRUE))
as.numeric(c('dog', 'cat', 'horse'))
as.numeric(c('7.1', '-3.5', '1.2'))
```

## Sample code
Convert data to integer values
```{r}
as.integer(c(31.2, 43.9, -22.76))
as.integer(c(FALSE, TRUE))
```

## Sample code
Nested converting
```{r}
as.logical(c('1.5','0.0','-3.2'))
as.logical(as.numeric(c('1.5','0.0','-3.2')))
```

## Sample code
Convert data to character values
```{r}
as.character(c(TRUE,FALSE,TRUE))
as.character(c(1.5, 2.2, 7.6))
as.character(c(5L, 22L))
```

## Sample code
Convert matrices to vectors
```{r}
mat <- matrix(1:9, nrow=3, ncol=3)
mat
as.vector(mat)
```

## Sample code
Convert vectors to a matrices
```{r}
vec <- c(1, 3, 5, 7, 9, 11)
vec
as.matrix(vec)
```

## Sample code
Convert vectors to a matrices, cont'd
```{r}
vec <- c(1, 3, 5, 7, 9, 11)
vec
matrix(vec, nrow=2, ncol=3)
```

## Sample code
Convert matrices to data frames
```{r}
mat <- matrix(1:12, ncol=6, nrow=2)
mat
as.data.frame(mat)
```

## Summary
- Use help() and several other functions to learn about functions and data sets
- Understand data types and structures
- Convert data types and structures
- Learn by practicing

## Thank you and see you next class


      







